Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2022-02-14 Manuscript/Tmatrix.tex
Cover letter

The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,vanAggelen_2014,Yang_2014a,Zhang_2015,Zhang_2016}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_2) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_1) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities (which should not appear in Coulombic systems with repulsive interactions only \cite{Scuseria_2013}), $\Nvir(\Nvir-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $\Nocc(\Nocc-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energy is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

@@

In terms of computational cost, the overall scaling of BSE@$GT$ is equivalent to BSE@$GW$ as they both correspond to seeking the lowest eigenvalues of a matrix of size $(2\Nocc\Nvir \times 2\Nocc\Nvir)$. 
Searching iteratively for the lowest eigenstates can be routinely performed via Davidson's algorithm with a $\order*{\Norb^4}$ computational cost. \cite{Davidson_1975}
The cost of the dynamical correction, which is thoroughly discussed in Ref.~\onlinecite{Loos_2020h}, is more expensive but is again equivalent in both formalisms.
The computational cost associated with the computation of the $T$-matrix and the screening $W$ both scale as $\order*{\Norb^6}$ as one must obtain all the eigenvalues and eigenvectors of the pp-RPA and ph-RPA problems, respectively.
However, the prefactor of the pp-RPA calculation is significantly larger than its ph-RPA counterpart due to the larger size of the pp-RPA matrices and its non-Hermitian nature. \cite{Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,Yang_2014a,Zhang_2015,Zhang_2016}
Moreover, within the $T$-matrix formalism, one must compute both the singlet and triplet contributions of the $T$-matrix, while for singlet states, only the singlet part of $W$ is required.
Although similar approaches remain to be developed for the $T$-matrix formalism, contour deformation and density fitting techniques can be efficiently implemented in the case of $GW$ to reduce the scaling to $\order*{\Norb^3}$. \cite{Duchemin_2019,Duchemin_2020,Duchemin_2021}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:ccl}
%%%%%%%%%%%%%%%%%%%%%%
We have derived and implemented, for the first time, the static and dynamic Bethe-Salpeter equations when one considers $T$-matrix quasiparticle energies as well as a $T$-matrix-based kernel.
The performance of the static scheme and its perturbative dynamical correction have been assessed by computing the neutral excited states of several molecular systems.
Our results suggest that the BSE@$GT$ formalism seems to be, overall, less accurate than BSE@$GW$ in the context of the computation of molecular excitation energies.
However, it still outperforms conventional methods such CIS and TDHF for singlet states.
It would be interesting to investigate its performance for the computation of ground-state correlation energies within the adiabatic connection fluctuation dissipation formalism where BSE@$GW$ has been shown to be particularly outstanding. \cite{Maggio_2016,Holzer_2018b,Loos_2020e}
The combination of $GT$ and $GW$ via the range separation of the Coulomb operator to avoid double counting of the low-order diagrams is also a promising avenue.
Work along these lines are currently under progress.
Finally, the unrestricted and spin-flip extensions of the present formalism are currently being developed.

>>>>

The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,vanAggelen_2014,Yang_2014a,Zhang_2015,Zhang_2016,Bannwarth_2020}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_2) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_1) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities (which should not appear in Coulombic systems with repulsive interactions only \cite{Scuseria_2013}), $\Nvir(\Nvir-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $\Nocc(\Nocc-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energy is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

@@

In terms of computational cost, the overall scaling of BSE@$GT$ is equivalent to BSE@$GW$ as they both correspond to seeking the lowest eigenvalues of a matrix of size $(2\Nocc\Nvir \times 2\Nocc\Nvir)$. 
Searching iteratively for the lowest eigenstates can be routinely performed via Davidson's algorithm with a $\order*{\Norb^4}$ computational cost. \cite{Davidson_1975}
The cost of the dynamical correction, which is thoroughly discussed in Ref.~\onlinecite{Loos_2020h}, is more expensive but is again equivalent in both formalisms.
The computational cost associated with the computation of the $T$-matrix and the screening $W$ both scale as $\order*{\Norb^6}$ in their standard implementation as one must obtain all the eigenvalues and eigenvectors of the pp-RPA and ph-RPA problems, respectively. \cite{Shenvi_2014} 
However, the prefactor of the pp-RPA calculation is significantly larger than its ph-RPA counterpart due to the larger size of the pp-RPA matrices and its non-Hermitian nature. \cite{Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,Yang_2014a,Zhang_2015,Zhang_2016}
Moreover, within the $T$-matrix formalism, one must compute both the singlet and triplet contributions of the $T$-matrix, while for singlet states, only the singlet part of $W$ is required.
Although similar approaches remain to be developed for the $T$-matrix formalism, contour deformation and density fitting techniques can be efficiently implemented in the case of $GW$ to reduce the scaling to $\order*{\Norb^3}$. \cite{Duchemin_2019,Duchemin_2020,Duchemin_2021}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:ccl}
%%%%%%%%%%%%%%%%%%%%%%
We have derived and implemented, for the first time, the static and dynamic Bethe-Salpeter equations when one considers $T$-matrix quasiparticle energies as well as a $T$-matrix-based kernel.
The performance of the static scheme and its perturbative dynamical correction have been assessed by computing the neutral excited states of several molecular systems.
Our results suggest that, in the context of the computation of molecular excitation energies, the BSE@$GT$ formalism performs best in few-electron systems where the electron density remains low.
For larger systems where the screening is more important, it seems to be, overall, less accurate than BSE@$GW$.
However, it still outperforms conventional methods such CIS and TDHF for singlet states.

It would be interesting to investigate its performance for the computation of ground-state correlation energies within the adiabatic connection fluctuation dissipation formalism where BSE@$GW$ has been shown to be particularly outstanding. \cite{Maggio_2016,Holzer_2018b,Loos_2020e}
The combination of $GT$ and $GW$ via the range separation of the Coulomb operator to avoid double counting of the low-order diagrams is also a promising avenue.
Work along these lines are currently under progress.
Finally, the unrestricted and spin-flip extensions of the present formalism are currently being developed.
